Deciphering enhancer rules underlying early human neuronal and neural crest development.

### 2025-01-24 manuscript.tex
result section 1 done.

\title{Title is a maximum of 145 characters, including spaces, and avoids the use of jargon and uncommon abbreviations}
\date{}

@@

\author[1,5,\orcidlink{0000-0000-0000-0000}]{First Last}
\author[1,2,5]{First Middle Last}
\author[2,3]{First M.M. Last}
\author[3,*]{First Last, Jr.}
\author[3,4,6,7,8,**]{F. Middle Last}

@@

\affil[1]{University A, London SW7 2AZ, UK}
\affil[2]{Department B, University B, Toronto, ON M5S 3H6, Canada}
\affil[3]{Division C, Department C, University C, Cambridge, MA, USA}

\affil[4]{Present address: 123 Sesame Street, New York, NY, USA}
\affil[5]{These authors contributed equally}
\affil[6]{Senior author}
\affil[7]{Twitter: @CorrespondingAuthor}
\affil[8]{Lead contact}

@@

\affil[*]{Correspondence: first.last.jr@university.edu}
\affil[**]{Correspondence: f.middle.last@university.edu}

@@

The summary (abstract) should consist of a single paragraph of 150 words or fewer.

@@

cell, press, LaTeX, template

@@

No subheadings, please. Use of abbreviations should be kept to a minimum, and nonstandard abbreviations should be defined when first used in the text.

@@

The results may be divided with subheadings. In our view, good subheadings convey information about the findings, so we encourage you to be specific. For example, "Factor X requires factor Y to function in process Z" is better than "Analysis of factors X and Y by approach Q." We recommend that you use similar language in your figure titles for clarity and structural harmony. Please do not use numbered subheadings. For some formats, ``results and discussion" may be combined under a single heading.

\subsection*{Figures}

For initial submissions, figures and their titles and legends can be run-in with the text. For final submissions, the image files must be uploaded as separate files, one complete file per figure. The figure titles and legends should be placed toward the end of the main text, after the declaration of interests. Our typesetters will place the figures and their titles and legends appropriately based on your in-text citations.

\subsection*{Citations}

To cite/link to display items (figures and tables) and/or sections of the manuscript, simply write, for example, ``(Figure 1)" or ``see discussion." Our team will do the rest. Please note that we do not number sections or subsections.

To cite references, you may use the cite command, e.g., ``Recent articles in \textit{Matter} and \textit{Cell} \cite{wang2020conformational,king2002,cates1984statics}  have shown ..." or  ``Many interesting discoveries have been reported, \cite{radzihovsky1995new,kapoor2022leakage,hantke2018my,alcock2006} which ..."

\subsection*{Equations}

Simple formulae should appear in line with the text whenever possible. You can write inline math by enclosing it between \verb|\(| and \verb|\)|, as in this example: \(x^2 + y^2 = z^2\). You can also enclose it between dollar signs (\texttt{\$}), as in this example: $E=mc^2$.

Larger, more complex formulae may appear on a new line, either by enclosing them between \verb|\[| and \verb|\]| or by using the \verb|displaymath| environment:

\[ x^n + y^n = z^n \]

\begin{displaymath}
\sqrt{x^2+1}
\end{displaymath}

If any equations or formulae need to be referred to or cited again later in the text, use the \verb|equation| environment to number them. Later, you can cite these as "Equation 1," "Equation 2," etc.

\begin{equation}
f(x) = \sum_{i=0}^{n} \frac{a_i}{1+x}
\end{equation}

@@

The discussion should explain the significance of the results and place them into a broader context. Subheadings are permitted.

@@

A "limitations" or "limitations of the study" subsection in the discussion is encouraged and may be required for some journals and some article formats.

@@

\noindent\includegraphics[width=0.85\linewidth]{Figure1.jpg}

The figure legend can be all one paragraph and describe the images (A), graphs (B), and plots (C), etc., together.
\newline
(A) Or each panel or group of panels can be described separately, as shown here and below.
\newline
(B) Graph of X, Y, and Z.
\newline
(C and D) If panels are grouped like this, please explicitly describe each panel, e.g., “Images showing SEM (C) and TEM (D).”
\newline
Please define all scale and error bars, and please review the Cell Press figure guidelines before submission: \href{https://www.cell.com/figureguidelines}{https://www.cell.com/figureguidelines}. Example figure created by Cassie Comeau, Cell Press.

>>>>

\title{Deciphering enhancer rules underlying early human neuronal and neural crest development.}
\date{}

@@

\author[1,2,3,\orcidlink{0000-0001-7907-1247}]{Seppe De Winter}
\author[1,2,3,*,\orcidlink{0000-0002-8006-0315}]{Stein Aerts}

@@

\affil[1]{VIB Center of AI \& computational Biology, Leuven, Belgium}
\affil[2]{VIB-KU Leuven Center for Brain and Disease Research, Leuven, Belgium}
\affil[3]{KU Leuven Department of Human Genetics, Leuven, Belgium}

@@

\affil[*]{Correspondence: stein.aerts@kuleuven.be}

@@

%%% The summary (abstract) should consist of a single paragraph of 150 words or fewer.

@@

@@

%%% No subheadings, please. Use of abbreviations should be kept to a minimum, and nonstandard abbreviations should be
%%% defined when first used in the text.

@@

\subsection*{Human neural tube organoids recapitulate early neuronal and neural crest development}
We made use of human neural tube organoids (hNTOs) to model early human neuronal and neural crest development. To this
end we embedded single cells from two induced pluripotent stem cell (iPSC) lines (Sigma and BJ1) and one embryonic stem
cell line (ESC; H9) into separate synthetic poly-ethylene glycol (PEG) based extracellular matrices. These were
incubated for 11 days and we supplemented the medium with retinoic acid (RA) and Smoothened agonist (SAG) from day 3
until day 5 (\textbf{Fig. 1a}). These conditions were previously shown to generate multiple neuronal progenitor and
neural crest states (REF). In order to characterize the organoid-derived cell types to a greater level of detail at both
transcriptomic as well as chromatin accessibility level we performed 10x single-cell multiomics (combined scATAC-seq and
scRNA-seq) at four time points of differentiation (day 4-5, day 6-7, day 8-9 and day 10-110; \textbf{Fig. 1a}),
resulting in 22,862 high quality cells (\textbf{Fig. 1b}). Note, that similar transcriptomic and chromatin accessibility
states were obtained from all three cell lines (\textbf{FIGURE SUPPLEMENT}) demonstrating the robustness of the
organoid system.\par
To evaluate the biological relevance of the cell types and states obtained from the organoid, we performed 10x single
cell multiome on the head of a single four weeks post conception human embryo, resulting in 25,785 high quality cells (
\textbf{Fig. 1c}). We annotated both the organoid and embryo cells based on the expression of marker genes
(\textbf{TABLE SUPPLEMENT}). In both systems we could identify neuronal progenitor cells, early differentiating neurons,
neural crest cells, facial mesenchyme cells and early differentiating peripheral neurons (\textbf{Fig. 1b-c}).
Additionally, in the organoid only we could identify a pluripotent stem cell population (\textbf{Fig. 1b}) and in the
embryo only we could identify an endothelial, perivascular macrophage (PVM) / microglia (MIC) and otic placode
population (\textbf{Fig. 1c}).\par
Next, we sub-clustered the neuronal progenitor, early neuron, and neural crest cell types (\textbf{Fig. 1d-f}). Both
organoid and embryo neuronal progenitors show a gradient of different cell states reflecting dorso-ventral patterning as
is evident from the expression of dorso-ventral marker TFs (REF): from ventral to dorsal ARX, NKX2-2, PAX6, IRX3, ZIC1 and
GDF7 (\textbf{Fig. 1d}). In both the organoid and embryo system we obtained multiple early neuronal differentiation states.
A cluster that still expresses the neuronal progenitor marker SOX2, cells that express the early neuronal induction
marker ASCL1 and distinct clusters marked by the expression of GAD1, UNCX and GATA2 (\textbf{Fig. 1e}). Finally,
focusing on the neural crest cells, in both the organoid and the embryo we obtained a population of pre-migratory neural
crest (GRHL2 expression and absence of ZEB2 expression), migratory neural crest (SOX10 and ZEB2 expression), facial
mesenchyme cells (TWIST1 expression) and early differentiation peripheral neurons (NEUROD1 expression). In the embryo,
but not in the organoid, we observed an additional population of cells marked by the expression of SOX10 and PAX2
reflecting the otic placode (\textbf{Fig. 1f}).\par
Given that we performed single-cell multiome on the entire head of the four weeks post conception embryo we also
assessed the expression of anterior-posterior marker TFs. While the organoid cells had either an anterior forebrain
(SIX3) or anterior spinal cord (HOXB3) identity, we observed a gradient of anterior-to-posterior identities in the
embryo cells marked by the expression of anterior-posterior marker TF (REF): from anterior to posterior SIX3, EMX2,
FOXG1, BARHL1, EN1 and HOXB3 (\textbf{Fig. 1g}).\par
Cell cycle arrest is an important hallmark of the start of neuronal differentiation (REF), for this reason we scored
both organoid and embryo cells for the expression of marker genes of the G1, G2M and S phase of the cell cycle and
quantified the fraction of cells in each phase stratified by the main cell types (\textbf{Fig. 1h}). From this it is
clear that both early differentiation peripheral neurons and to a greater extend early differentiating neurons of the
central nervous system are in cell cycle arrest, while the other cell types are still cycling (\textbf{Fig. 1h}).\par
Finally, to assess whether the organoid-derived cell types and states are also similar to the embryo-derived cell types
and states on a chromatin accessibility level we identified sets of co-accessible regions in the organoid model using
topic modeling. We focused on topics (i.e., sets of co-accessible regions) that were enriched for marker TF motifs and
calculated the average expression of marker genes for each cell group where these regions were co-accessible, for both
organoid and embryo cells. We observed similar patterns of marker gene expression for the same cell state (defined by
the topic) across the organoid and embryo (\textbf{Fig. 1i}) with an average Pearson correlation coefficient of \textbf{X.X}
across identical states (\textbf{FIGURE SUPPLEMENT}).

\noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_1.png}

\textbf{a}, Schematic. \textbf{b-c}, UMAP based on gene expression profiles of organoid (\textbf{b}) and embryo
(\textbf{c}) cells colored by cell type identify. \textbf{d-f}, UMAP based on gene expression profiles of organoid (top)
and embryo (bottom) cells types showing sub-clusters of neuronal progenitors (\textbf{d}), early differentiating neurons
(\textbf{e}) and neural crest (\textbf{f}) along with expression of indicated genes in red-green-blue color scale.
\textbf{g}, UMAP based on gene expression profiles of organoid (top; UMAP axis 1 and 2 shown) and embryo (bottom; UMAP
axis 2 and 3 shown) for neuronal progenitor subcluster along with the expression of indicated genes in red-green-blue
color scale. \textbf{h}, UMAP based on gene expression profiles in organoid (top left) and embryo (top right) colored by
cell cycle phase identity and quantification of percentage of each cell for each cell cycle phase stratified by cell
type in the organoid (bottom left) and embryo (bottom right). \textbf{i}, Average gene expression level in
$log_{2}(CPM)$
(dot color) and fraction of cells expressing the gene (dot size) per cell state, defined by topic modeling, for organoid
cells (top) and embryo cells (middle) and motif enrichment level as normalized enrichment score (NES; dot color) and
number of regions enriched for the motif (dot size) per topic (x-axis) and motif (y-axis; bottom). PVM/MIC: perivascular
microphage/microglia.

\subsection*{
  DeepNeuralTube: a sequence-to-function model revealing the enhancer code underlying early neuronal and neural crest 
  development
}

\noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_2.png}

\subsection*{Figure 2. }

@@

%%% The discussion should explain the significance of the results and place them into a broader context.
%%% Subheadings are permitted.

@@

%%% A "limitations" or "limitations of the study" subsection in the discussion is encouraged and may be required for some
%%% journals and some article formats.

@@
